---
title: "ordinal logistic regression practical"
author: "Azmi"
date: "22/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# environment

```{r}
library(pacman)
p_load(tidyverse, ordinal, summarytools, knitr, broom)
```

# Tutorial 1

## dataset

```{r}
dataset <- wine
head(dataset)
descr(dataset)
freq(dataset)
```

### ordinal::wine dataset

response - scoring of wine bitterness (numerical)
rating - ordered factors of response in 5 level (categorical - ordinal)
temp - temperature - cold vs warm (categorical)
contact - no vs yes (categorical)
bottle - 1 - 8 (categorical)
judge - 1 - 9 (categorical)

## short practical

analysis using proportional odds model or cummulative logit model

```{r}
levels(dataset$rating)
olrmod1 <- clm(rating ~ temp + contact, data = dataset)
summary(olrmod1)
tidy(olrmod1, conf.int = T)
tidy(olrmod1, conf.int = T, exponentiate = T)
```

note:

- beta coefficient is contant but alpha change for each pair of comparison
- however alpha (intercept) was not reported here. only SAS report intercept value
- the comparison is compared to bigger level

###fitted data

```{r}
fitprob.olrmod1 <- augment(olrmod1, conf.int = T)
head(fitprob.olrmod1)
fitcat.olrmod1 <- augment(olrmod1, type.predict = "class")
head(fitcat.olrmod1)
```


```{r}
crostab.olrmod1 <- table(fitcat.olrmod1$rating, fitcat.olrmod1$.fitted)
chisq.test(fitcat.olrmod1$rating, fitcat.olrmod1$.fitted)
```

###prediction

```{r}
newdata <- expand.grid(temp = levels(dataset$temp), contact = levels(dataset$contact))
fitprob.newdata <- augment(olrmod1, newdata = newdata, type.predict= "prob") ##same with predict(..., type = "prob")
fitclass.newdata <- augment(olrmod1, newdata = newdata, type.predict= "class") ##same with predict(..., type = "class")
```

- LP
note: no augment function for lp
```{r}
lp.newdata <- predict(olrmod1, newdata = newdata, type = "linear.predictor")
cbind(newdata, lp.newdata)
```

LP - use eta1.x value
eta2.x value??

### lp using newdata

```{r}
newdata
lp.newdata
coef_olrmod1 <- coef(olrmod1)
coef_olrmod1

lp_olrmod1_betax <- coef_olrmod1[5]*0 + coef_olrmod1[6]*0 
lp_olrmod1_betax
```


### proportional odds assumption checking

- check for proportional odds for variable contact, and compare with model that have both temp and contact

```{r}
olrmod1_connom <- clm(rating~temp, nominal = ~contact, data = dataset)
anova(olrmod1, olrmod1_connom)

olrmod1_tempnom <- clm(rating~contact, nominal = ~temp, data = dataset)
anova(olrmod1, olrmod1_tempnom)

```

- p = 0.904 --> not significant --> assumption met (proportionality does present for variable contact)

ANOVA here is compare loglikelihood



## another package --> MASS

```{r}
library(MASS)
wine_polr <- polr(rating ~ temp + contact, data = dataset)
wine_polr
summary(wine_polr)
```

predict probability
```{r}
prob_winepolr <- augment(wine_polr, type.predict = "probs")
prob_winepolr
```



# Tutorial 2

## dataset

```{r}
library(haven)
lowbwt <- read_dta("lowbwt.dta")
lowbwt <- lowbwt %>% mutate_if(is.labelled, funs(as_factor(.)))
head(lowbwt)
descr(lowbwt)
freq(lowbwt)
lowbwt1 <- lowbwt %>% 
  select(smoke, weight, lwt) %>% 
  mutate(bwtcat = cut(weight,
                      breaks = c(0,2500,3000,3500, 5000),
                      labels = c("<= 2500", "2501 - 3000", "3001 - 3500", "> 3500")))
lev <- c('>3500','3001-3500','2501-3000','=<2500')
```


Variables

- lbw: low birth weight
  - 0 - >= 2500g (no)
  - 1 - < 2500g (yes)
- age: age of mother
- lwt: weight mohter at last menstrual period (pounds)
- race: 
  - 1 - white
  - 2 - black
  - 3 - other
- smoke: smoking status during pregnancy
  - 0 - no
  - 1 - yes
- history of prem labour
  - 0 - none
  - 1 - one
  - 2 - two, etc
- hyper: history of hpt
  - 0 - no
  - 1 - yes
- urirr: uterine irritability
  - 0 - no
  - 1 - yes
- pvft: no. of physician visit during 1st trimester
  - 0 - none
  - 1 - 1
  - 2 - 2, etc
- weight: birth weight


```{r}
freq(lowbwt1$bwtcat)
```

```{r}
lowbwt1$bwtcata <- ordered(lowbwt1$bwtcat, levels=c("> 3500", "3001 - 3500", "2501 - 3000", "<= 2500"))
freq(lowbwt1$bwtcata)
freq(lowbwt1$bwtcat)
```

## adjacent model

```{r}
library(VGAM)
lowbwt1a <- lowbwt1a %>%
  mutate(bwtnoord = fct_relevel(bwtcat, lev))
levels(lowbwt1a$bwtnoord)
ajm_bwt <- vglm(bwtnoord ~ smoke, multinomial, data=lowbwt1a)
summary(ajm_bwt)
```


## multinomial or baseline logit model

baseline logit model - multiple binomial - compare to lower group

### g1 (> 3500) vs g2 (3001 - 3500)

```{r}
lowbwt1a <- lowbwt1[lowbwt1$bwtcata == "> 3500"|lowbwt1$bwtcata == "3001 - 3500",]
freq(lowbwt1a$bwtcata)
blm1 <- glm(bwtcata ~ smoke, family = binomial, data = lowbwt1a)
blm1
summary(blm1)
```

### g1 (> 3500) vs g3 (2501 - 3000)

```{r}
lowbwt1b <- lowbwt1[lowbwt1$bwtcata == "> 3500"|lowbwt1$bwtcata == "2501 - 3000",]
freq(lowbwt1b$bwtcata)
blm2 <- glm(bwtcata ~ smoke, family = binomial, data = lowbwt1b)
blm2
summary(blm2)
```


### g1 (> 3500) vs g4 (<= 2500)

```{r}
lowbwt1c <- lowbwt1[lowbwt1$bwtcata == "> 3500"|lowbwt1$bwtcata == "<= 2500",]
freq(lowbwt1c$bwtcata)
blm3 <- glm(bwtcata ~ smoke, family = binomial, data = lowbwt1c)
blm3
summary(blm3)
```

## continuation ratio

### g1 vs g2

```{r}
cr1 <- blm1
cr1
summary(cr1)
```

### g1 + g2 vs g3

```{r}
lowbwt2a <- lowbwt1[lowbwt1$bwtcata=="> 3500"|lowbwt1$bwtcata=="3001 - 3500"|lowbwt1$bwtcata=="2501 - 3000",]
freq(lowbwt2a)
lowbwt2a$bwtcata1 <- ifelse(lowbwt2a$bwtcata=="> 3500", 0,
                           ifelse(lowbwt2a$bwtcata=="3001 - 3500", 0,1))
freq(lowbwt2a$bwtcata1)
cr2 <- glm(bwtcata1 ~ smoke, family = binomial, data = lowbwt2a)
cr2
summary(cr2)
```

### g1+g2+g3 vs g4

```{r}
lowbwt2b <- lowbwt1
freq(lowbwt2b)
lowbwt2b$bwtcata1 <- ifelse(lowbwt2b$bwtcata=="> 3500", 0,
                           ifelse(lowbwt2b$bwtcata=="3001 - 3500", 0,
                                  ifelse(lowbwt2b$bwtcata=="2501 - 3000", 0,1)))
freq(lowbwt2b$bwtcata1)
cr3 <- glm(bwtcata1 ~ smoke, family = binomial, data = lowbwt2b)
cr3
summary(cr3)
```


# Tutorial 3

## cumulative logit models = proportional odds model

```{r}
levels(lowbwt1$bwtcat)
descr(lowbwt1$weight)
```

```{r}
levels(lowbwt1$bwtcat)
clmmod1 <- clm(bwtcat ~ lwt, data = lowbwt1)
clmmod1
summary(clmmod1)
```

```{r}
tidy(clmmod1, conf.int = T)
tidy(clmmod1, conf.int = T, exponentiate = T)
```

### fitted data - probability

```{r}
fitted_clm <- augment(clmmod1)
fitted_clm
```



## plor package - Proportional Odds Linear Regression

```{r}
library(MASS)
polr_cr3 <- polr(bwtcat ~ lwt, data = lowbwt1, Hess = T)
polr_cr3
summary(polr_cr3)
```

```{r}
tidy(polr_cr3)
tidy(polr_cr3, exponentiate = T)
```

### proportional odds assumption

```{r}
library(brant)
brant(polr_cr3)
```

### fitted data - probability

```{r}
fitted_polr_cr3 <- predict(polr_cr3, type = "p")
head(fitted_polr_cr3)
```