---
title: "ordinal logistic regression practical"
author: "Azmi"
date: "22/02/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# environment

```{r}
library(pacman)
p_load(tidyverse, ordinal, summarytools, knitr, broom)
```

# Tutorial 1

## dataset

```{r}
dataset <- wine
head(dataset)
descr(dataset)
freq(dataset)
```

### ordinal::wine dataset

response - scoring of wine bitterness (numerical)
rating - ordered factors of response in 5 level (categorical - ordinal)
temp - temperature - cold vs warm (categorical)
contact - no vs yes (categorical)
bottle - 1 - 8 (categorical)
judge - 1 - 9 (categorical)

## short practical

```{r}
levels(dataset$rating)
olrmod1 <- clm(rating ~ temp + contact, data = dataset)
summary(olrmod1)
tidy(olrmod1, conf.int = T)
tidy(olrmod1, conf.int = T, exponentiate = T)
```

note:

- beta coefficient is contant but alpha (threshold coefficient) change for each pair of comparison
- the comparison is compared to bigger level

###fitted data

```{r}
fitprob.olrmod1 <- augment(olrmod1, conf.int = T)
head(fitprob.olrmod1)
fitcat.olrmod1 <- augment(olrmod1, type.predict = "class")
head(fitcat.olrmod1)
```


```{r}
crostab.olrmod1 <- table(fitcat.olrmod1$rating, fitcat.olrmod1$.fitted)
chisq.test(fitcat.olrmod1$rating, fitcat.olrmod1$.fitted)
```

###prediction

```{r}
newdata <- expand.grid(temp = levels(dataset$temp), contact = levels(dataset$contact))
fitprob.newdata <- augment(olrmod1, newdata = newdata, type.predict= "prob") ##same with predict(..., type = "prob")
fitclass.newdata <- augment(olrmod1, newdata = newdata, type.predict= "class") ##same with predict(..., type = "class")
```

- LP
note: no augment function for lp
```{r}
lp.newdata <- predict(olrmod1, newdata = newdata, type = "linear.predictor")
cbind(newdata, lp.newdata)
```

### proportional odds assumption checking

- check for proportional odds for variable contact, and compare with model that have both temp and contact

```{r}
olrmod1_nom <- clm(rating~temp, nominal = ~contact, data = dataset)
anova(olrmod1,olrmod1_nom)
```

- p = 0.904 --> not significant --> assumption met (proportionality does present for variable contact)


## another package --> MASS

# Tutorial 2

## dataset

```{r}
library(haven)
lowbwt <- read_dta("lowbwt.dta")
head(lowbwt)
```

Variables

- lbw: low birth weight
  - 0 - >= 2500g (no)
  - 1 - < 2500g (yes)
- age: age of mother
- lwt: weight mohter at last menstrual period (pounds)
- race: 
  - 1 - white
  - 2 - black
  - 3 - other
- smoke: smoking status during pregnancy
  - 0 - no
  - 1 - yes
- history of prem labour
  - 0 - none
  - 1 - one
  - 2 - two, etc
- hyper: history of hpt
  - 0 - no
  - 1 - yes
- urirr: uterine irritability
  - 0 - no
  - 1 - yes
- pvft: no. of physician visit during 1st trimester
  - 0 - none
  - 1 - 1
  - 2 - 2, etc
  
  
